---
title: "Sleep Disorder"
format: html
---

## Sleep Health and Lifestyle Analysis Project

By:

1.  Amabel Nabila (455854)

    -   Data Visualization and Analysis

2.  Mustafa Ceyhun Vural (455158)

    -   

3.  Onur Durmus (455860)

    -   

4.  Zehra Usta (455847)

    -   

## Project Objective

This project serves as the final assignment for the Reproducible Research course. It utilizes a dataset sourced from Kaggle and involves translating the code from Python to R. The aim of this project is to analyze the sleep health of individuals based on their respective lifestyles. Variables in this project include:

-   Gender

-   Age

-   Occupation

-   Sleep Duration

-   Quality of Sleep

-   Physical Activity Level

-   Stress Level

-   BMI Category

-   Blood Pressure

-   Heart Rate

-   Daily Steps

-   Sleep Disorder

## Introduction

Sleep disorders (or sleep-wake disorders) are difficulties with the quality, timing, and quantity of sleep that cause daily discomfort and impairment in functioning. Sleep-wake disturbances are frequently associated with physical or mental health illnesses such as depression, anxiety, or cognitive difficulties. There are various forms of sleep-wake disorders, with insomnia being the most frequent. Other sleep-wake problems include sleep apnea. Insomnia is a common sleep problem that makes it difficult to get asleep and remain asleep. Meanwhile, sleep apnea is a condition in which you stop breathing while asleep.

## Library

```{r}
# Libraries used
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(corrplot)
```

## Data

```{r}
# Read the data
data <- read.csv('Sleep_health_and_lifestyle_dataset.csv')
```

## Data Cleaning

```{r}
# Drop NA (if existed)
data <- na.omit(data)

# Standardize "Normal Weight" and "normal" to "Normal" in "BMI Category"
data <- data %>%
  mutate(BMI.Category = ifelse(BMI.Category %in% c("Normal Weight", "normal"), "Normal", BMI.Category))

# View data cleaned
data
```

```{r}
# Split the Blood.Pressure column into two new columns and convert to numeric
data <- data %>%
  separate(col = Blood.Pressure, into = c("Systolic Pressure", "Diastolic Pressure"), sep = "/", convert = TRUE) %>%
  mutate(`Systolic Pressure` = as.numeric(`Systolic Pressure`),
         `Diastolic Pressure` = as.numeric(`Diastolic Pressure`))

# Check the structure of the updated data frame
str(data)
```

```{r}
# Define a function to categorize blood pressure
categorize_blood_pressure <- function(systolic, diastolic) {
  if (is.na(systolic) | is.na(diastolic)) {
    return(NA)
  } else if (systolic < 120 & diastolic < 80) {
    return('Optimal')
  } else if (systolic <= 139 | diastolic <= 89) {
    return('Normal')
  } else if (systolic >= 140 | diastolic >= 90) {
    return('Hypertension')
  } else {
    return('Other')
  }
}

# Apply the categorization function to create the Blood Pressure Category column
data <- data %>%
  rowwise() %>%
  mutate(`Blood.Pressure.Category` = categorize_blood_pressure(`Systolic Pressure`, `Diastolic Pressure`))

# View the newest data
data
```

## Data Visualization

```{r}
# Create a box plot for Age distribution
age_boxplot <- ggplot(data, aes(x = '', y = Age)) + 
  geom_boxplot() +
  labs(title = "Age Distribution", y = "Age", x = "") +
  theme_minimal()

# Create a histogram for Age distribution
age_histogram <- ggplot(data, aes(x = Age)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()

# Create a box plot for Quality of Sleep distribution
quality_of_sleep_boxplot <- ggplot(data, aes(x = '', y = Quality.of.Sleep)) + 
  geom_boxplot() +
  labs(title = "Quality of Sleep Distribution", y = "Quality of Sleep", x = "") +
  theme_minimal()

# Create a histogram for Quality of Sleep distribution
quality_of_sleep_histogram <- ggplot(data, aes(x = Quality.of.Sleep)) + 
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") +
  labs(title = "Quality of Sleep Distribution", x = "Quality of Sleep", y = "Count") +
  theme_minimal()

# Create a box plot for Physical Activity Level distribution
physical_activity_boxplot <- ggplot(data, aes(x = '', y = Physical.Activity.Level)) + 
  geom_boxplot() +
  labs(title = "Physical Activity Level Distribution", y = "Physical Activity Level", x = "") +
  theme_minimal()

# Create a histogram for Physical Activity Level distribution
physical_activity_histogram <- ggplot(data, aes(x = Physical.Activity.Level)) + 
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Physical Activity Level Distribution", x = "Physical Activity Level", y = "Count") +
  theme_minimal()

# Create a box plot for Stress Level distribution
stress_level_boxplot <- ggplot(data, aes(x = '', y = Stress.Level)) + 
  geom_boxplot() +
  labs(title = "Stress Level Distribution", y = "Stress Level", x = "") +
  theme_minimal()

# Create a histogram for Stress Level distribution
stress_level_histogram <- ggplot(data, aes(x = Stress.Level)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Stress Level Distribution", x = "Stress Level", y = "Count") +
  theme_minimal()

# Create a box plot for Heart Rate distribution
heart_rate_boxplot <- ggplot(data, aes(x = '', y = Heart.Rate)) + 
  geom_boxplot() +
  labs(title = "Heart Rate Distribution", y = "Heart Rate", x = "") +
  theme_minimal()

# Create a histogram for Heart Rate distribution
heart_rate_histogram <- ggplot(data, aes(x = Heart.Rate)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Heart Rate Distribution", x = "Heart Rate", y = "Count") +
  theme_minimal()

# Create a box plot for Daily Steps distribution
daily_steps_boxplot <- ggplot(data, aes(x = '', y = Daily.Steps)) + 
  geom_boxplot() +
  labs(title = "Daily Steps Distribution", y = "Daily Steps", x = "") +
  theme_minimal()

# Create a histogram for Daily Steps distribution
daily_steps_histogram <- ggplot(data, aes(x = Daily.Steps)) + 
  geom_histogram(binwidth = 500, fill = "blue", color = "black") +
  labs(title = "Daily Steps Distribution", x = "Daily Steps", y = "Count") +
  theme_minimal()

# Combine the plots into a single figure
combined_plots <- (age_boxplot | age_histogram) / 
  (quality_of_sleep_boxplot | quality_of_sleep_histogram) / 
  (physical_activity_boxplot | physical_activity_histogram) / 
  (stress_level_boxplot | stress_level_histogram) / 
  (heart_rate_boxplot | heart_rate_histogram) / 
  (daily_steps_boxplot | daily_steps_histogram) 

# Define the layout of the combined plot
combined_plot_layout <- combined_plots + 
  plot_layout(ncol = 2)

# Print the combined plot with the layout
print(combined_plot_layout)
```

**Age Distribution**:

The box plot indicates a fairly uniform age distribution, with a few outliers on the higher end. The histogram verifies this, displaying a fairly consistent distribution across age groups with a little rise in frequency at the middle age range.

**Quality of Sleep Distribution**:

The box plot shows that most values are concentrated in the center of the scale, implying that the majority of people have intermediate sleep quality, with occasional outliers indicating very good sleep quality. The histogram confirms this, demonstrating that most people report sleep quality in the center of the scale, with fewer reporting extremely high or terrible sleep quality.

**Physical Activity Level Distribution:**

The box plot for physical activity levels shows a broad range of values, with outliers at both the low and high ends. The histogram has a bimodal distribution, with peaks at both the low and high ends of the activity scale. This might imply that there are two separate groups in the population: one that is less active and one that is more active.

**Stress Level Distribution**:

According to the box plot, stress levels are typically close to the median, with no extreme outliers. The histogram shows a rather uniform distribution, with a tiny peak at higher stress levels.

**Heart Rate Distribution:**

The heart rate box plot displays a reasonably compact interquartile range, indicating that the majority of people have a resting heart rate within a restricted range, despite the presence of multiple outliers. The histogram shows a normal distribution with a peak at 70 beats per minute, which is typical for a resting heart rate in adults.

**Daily Steps Distribution**:

The box plot depicts a wide distribution, reflecting variation in the number of daily steps across individuals, with a few outliers suggesting extremely high step counts. The histogram shows that the majority of people have a daily step count on the low end of the scale, with fewer people reaching higher step counts.\

```{r}
# Calculate mean, median, and mode
mean_sleep <- mean(data$Quality.of.Sleep, na.rm = TRUE)
median_sleep <- median(data$Quality.of.Sleep, na.rm = TRUE)
mode_sleep <- as.numeric(names(sort(table(data$Quality.of.Sleep), decreasing = TRUE)[1]))

# Create the histogram and add lines for mean, median, and mode
quality_of_sleep_plot <- ggplot(data, aes(x = Quality.of.Sleep)) +
  geom_histogram(aes(y = ..count..), binwidth = 1, fill = "lightblue", color = "black") +
  geom_vline(aes(xintercept = mean_sleep, color = "Mean"), linetype = "dashed", linewidth = 1) +
  geom_vline(aes(xintercept = median_sleep, color = "Median"), linetype = "dotted", linewidth = 1) +
  geom_vline(aes(xintercept = mode_sleep, color = "Mode"), linetype = "dotdash", linewidth = 1) +
  scale_color_manual(name = "Statistics", values = c("Mean" = "red", "Median" = "green", "Mode" = "blue")) +
  labs(title = "Quality of Sleep Distribution", x = "Quality of Sleep", y = "Count") +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(vjust = -0.2),
    axis.title.y = element_text(vjust = 1.2)
  )

# Print the plot
print(quality_of_sleep_plot)
```

The majority of participants reported excellent quality of sleep (average scores of 7 and 8). The fact that the mode is higher than both the mean and the median indicates that, while some people may have poor sleep quality, this group is more likely to have good sleep.\

```{r}
# Column names
quality_of_sleep_col <- "Quality.of.Sleep"
grouping_cols <- c("Gender", "Occupation", "BMI.Category", "Blood.Pressure.Category", "Sleep.Disorder")

plots <- list()


for(grouping_col in grouping_cols) {
  # Check if each group has at least two data points
  valid_groups <- data %>%
    group_by(!!sym(grouping_col)) %>%
    filter(n() > 1) %>%
    pull(!!sym(grouping_col)) %>%
    unique()
  
  # Create the density plot only for valid groups
  if (length(valid_groups) > 0) {
    p <- ggplot(data %>% filter(!!sym(grouping_col) %in% valid_groups), 
                aes_string(x = quality_of_sleep_col, fill = grouping_col)) +
      geom_density(alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Quality of Sleep by", grouping_col))
    plots[[grouping_col]] <- p
  } else {
    warning(paste("Not enough data to plot", grouping_col))
  }
}

# Check your list of plots
plots

# To display the plots
if (length(plots) > 0) {
  library(gridExtra)
  do.call(grid.arrange, c(plots, ncol = 2))
}
```

The investigation reveals that there are distinct patterns and variances in sleep quality based on gender, employment, BMI, blood pressure, and the prevalence of sleeping disorders. Lifestyle and health circumstances appear to have a substantial impact on sleep quality. \

```{r}
# Boxplot for Age by Gender
age_plot <- ggplot(data, aes(x = Gender, y = Age, fill = Gender)) +
  geom_boxplot() +
  labs(title = "1. How old are they?", y = "Age", x = "Gender") +
  theme_minimal()

# Boxplot for Sleep Duration by Gender
sleep_duration_plot <- ggplot(data, aes(x = Gender, y = `Sleep.Duration`, fill = Gender)) +
  geom_boxplot() +
  labs(title = "2. How long do they sleep?", y = "Sleep.Duration", x = "Gender") +
  theme_minimal()

# Boxplot for Quality of Sleep by Gender
quality_sleep_plot <- ggplot(data, aes(x = Gender, y = `Quality.of.Sleep`, fill = Gender)) +
  geom_boxplot() +
  labs(title = "3. How well do they sleep?", y = "Quality of Sleep", x = "Gender") +
  theme_minimal()

# Combine the plots into a single plot window
library(gridExtra)
grid.arrange(age_plot, sleep_duration_plot, quality_sleep_plot, ncol = 3)
```

**Age:**

The age boxplot suggests that, on average, women in this dataset are older than men, as indicated by the higher median line for women in the box. Comparable ranges for the top and bottom of the box and the interquartile range (IQR, the height of the box) point to similar age variability for both genders. There are men and females that are both younger and older than the core age categories, according to the whiskers, which depict the dispersion of the remaining data. However, the range for females is larger, indicating greater ages.

**Sleep Duration:**

The boxplot illustrates that men have a somewhat shorter median sleep time than women. The identical IQRs for the two genders suggest that the middle 50% of people have similar sleep duration distributions. Males appear to have a significantly larger distribution of sleep time than females, as indicated by the slightly greater range for males.

**Quality of Sleep:**

The median line suggests that when it comes to sleep quality, women appear to report a little greater median quality of sleep than men. For both genders, the dispersion of values (IQR) is comparable, though. Females tend to have a greater general range than males, suggesting that the quality of their sleep is more variable, with some reporting extremely good and some very low quality.

```{r}
# Count the number of participants in each occupation
occupation_count <- data %>%
  group_by(Occupation) %>%
  summarise(Participants = n_distinct(Person.ID))

# Calculate the average quality of sleep for each occupation
avg_quality_of_sleep <- data %>%
  group_by(Occupation) %>%
  summarise(AvgQualityOfSleep = mean(Quality.of.Sleep, na.rm = TRUE))

# Number of Participants by Occupation
occupation_count_plot <- ggplot(occupation_count, aes(x = reorder(Occupation, -Participants), y = Participants)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "1. What are their occupations?", x = "Number of participants", y = "Occupation") +
  theme_minimal()

# Average Quality of Sleep by Occupation
quality_sleep_plot <- ggplot(avg_quality_of_sleep, aes(x = reorder(Occupation, AvgQualityOfSleep), y = AvgQualityOfSleep)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "2. Who sleeps better, and who worse?", x = "Avg Quality of Sleep", y = "Occupation") +
  theme_minimal()

# Print the plots
print(occupation_count_plot)
print(quality_sleep_plot)
```

**Occupational Distribution**:

The distribution of participants among different vocations is seen in the first chart. The most common occupation is nursing, which is followed by engineering and medicine. In the sample, managers and sales reps are the least represented. This distribution may represent the true demographic composition of the pool from which participants were selected, or it may show the interest or availability of various occupations in taking part in the study.\

**Average Quality of Sleep by Occupation**:

The second graph displays the average sleep quality for various professions. Engineers seem to have the greatest average sleep quality, followed by accountants and attorneys. Conversely, scientists and sales people report the lowest average quality of sleep, which might be attributed to issues relating to their line of work or greater levels of stress related to their jobs. Despite the important importance of their work, doctors do not necessarily get better sleep than other professions, as seen by the fact that, despite their large representation in the dataset, their average sleep quality does not top the list.\

```{r}
# Count of participants by Sleep Disorder
sleep_disorder_count <- ggplot(data, aes(x = Sleep.Disorder)) +
  geom_bar(fill = "skyblue") +
  labs(title = "1. How many suffer from Sleep Disorders?", x = "Sleep Disorder", y = "Number of Participants") +
  theme_minimal()

# Average Quality of Sleep by Sleep Disorder
avg_sleep_quality_by_disorder <- ggplot(data, aes(x = Sleep.Disorder, y = Quality.of.Sleep)) +
  geom_bar(stat = "summary", fun = "mean", fill = "skyblue") +
  labs(title = "2. Who sleeps better?", x = "Sleep Disorder", y = "Avg Sleep Quality") +
  theme_minimal()

# Sleep quality boxplot by Sleep Disorder
sleep_quality_boxplot <- ggplot(data, aes(x = Sleep.Disorder, y = Quality.of.Sleep)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "3. Sleep quality boxplot by Sleep Disorders", x = "Sleep.Disorder", y = "Sleep.Quality") +
  theme_minimal()

# Print the plots
print(sleep_disorder_count)
print(avg_sleep_quality_by_disorder)
print(sleep_quality_boxplot)
```

```{r}
# BMI Data Plotting
# Count the number of participants in each BMI category
bmi_counts <- data %>%
  group_by(BMI.Category) %>%
  summarise(Number.of.Participants = n())

# Calculate the average quality of sleep for each BMI category
bmi_quality <- data %>%
  group_by(BMI.Category) %>%
  summarise(Avg.Quality.of.Sleep = mean(Quality.of.Sleep, na.rm = TRUE))

# Bar plot for BMI category participant count
bmi_count_plot <- ggplot(bmi_counts, aes(x = BMI.Category, y = Number.of.Participants, fill = BMI.Category)) +
  geom_bar(stat = "identity") +
  labs(title = "1. What is your BMI category?", x = "BMI Category", y = "Number of participants") +
  theme_minimal()

# Bar plot for average quality of sleep by BMI category
bmi_quality_plot <- ggplot(bmi_quality, aes(x = BMI.Category, y = Avg.Quality.of.Sleep, fill = BMI.Category)) +
  geom_bar(stat = "identity") +
  labs(title = "2. Who sleeps better?", x = "BMI Category", y = "Avg Quality of Sleep") +
  theme_minimal()

# Boxplot for quality of sleep by BMI category
bmi_quality_boxplot <- ggplot(data, aes(x = BMI.Category, y = Quality.of.Sleep, fill = BMI.Category)) +
  geom_boxplot() +
  labs(title = "3. Sleep quality boxplot by BMI Category", x = "BMI Category", y = "Quality of Sleep") +
  theme_minimal()

# Display the plots
bmi_count_plot
bmi_quality_plot
bmi_quality_boxplot
```

```{r}
# Blood Pressure Data Plotting
# Repeat the same process for blood pressure data
bp_counts <- data %>%
  group_by(Blood.Pressure.Category) %>%
  summarise(Number.of.Participants = n())

bp_quality <- data %>%
  group_by(Blood.Pressure.Category) %>%
  summarise(Avg.Quality.of.Sleep = mean(Quality.of.Sleep, na.rm = TRUE))

bp_count_plot <- ggplot(bp_counts, aes(x = Blood.Pressure.Category, y = Number.of.Participants, fill = Blood.Pressure.Category)) +
  geom_bar(stat = "identity") +
  labs(title = "1. How is your Blood Pressure level?", x = "Blood Pressure Category", y = "Number of participants") +
  theme_minimal()

bp_quality_plot <- ggplot(bp_quality, aes(x = Blood.Pressure.Category, y = Avg.Quality.of.Sleep, fill = Blood.Pressure.Category)) +
  geom_bar(stat = "identity") +
  labs(title = "2. Who sleeps better?", x = "Blood Pressure Category", y = "Avg Quality of Sleep") +
  theme_minimal()

bp_quality_boxplot <- ggplot(data, aes(x = Blood.Pressure.Category, y = Quality.of.Sleep, fill = Blood.Pressure.Category)) +
  geom_boxplot() +
  labs(title = "3. Sleep Quality boxplot by Blood Pressure Category", x = "Blood Pressure Category", y = "Quality of Sleep") +
  theme_minimal()

# Display the plot
bp_count_plot
bp_quality_plot
bp_quality_boxplot
```

```{r}
# Select only the numeric columns from the data
numeric_data <- data[sapply(data, is.numeric)]

# Calculating the correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Create the correlation heatmap
corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45,
         diag = FALSE)  

```

```{r}
# Scatter plot of Sleep Duration vs Quality of Sleep
plot1 <- ggplot(data, aes(x = Quality.of.Sleep, y = Sleep.Duration)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "Long Rest = Good Rest",
       x = "Quality of Sleep",
       y = "Sleep Duration (hours)")

# Scatter plot of Stress Level vs Quality of Sleep
plot2 <- ggplot(data, aes(x = Quality.of.Sleep, y = Stress.Level)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "More Stress = Bad Rest",
       x = "Quality of Sleep",
       y = "Stress Level")

#Show the plots
plot1
plot2
```

```{r}
# Scatter plot of Physical Activity Level vs Sleep Quality
plot3 <- ggplot(data, aes(x = Quality.of.Sleep, y = Physical.Activity.Level)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "A: Sleep quality based on physical activity",
       x = "Sleep Quality",
       y = "Physical Activity Level (minutes)")

# Scatter plot of Age vs Sleep Duration
plot4 <- ggplot(data, aes(x = Age, y = Sleep.Duration)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "B: The older you get, the less you sleep",
       x = "Age",
       y = "Sleep Duration (hours)")

# Scatter plot of Heart Rate vs Sleep Duration
plot5 <- ggplot(data, aes(x = Heart.Rate, y = Sleep.Duration)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "C: Quickest Heartbeat, Fewest Sleep",
       x = "Heart Rate",
       y = "Sleep Duration (hours)")

# Scatter plot of Physical Activity vs Sleep Duration
plot6 <- ggplot(data, aes(x = Physical.Activity.Level, y = Sleep.Duration)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "D: Engaging in physical exercise can facilitate sleep",
       x = "Physical Activity",
       y = "Sleep Duration (hours)")

#Show the plots
plot3
plot4
plot5
plot6
```

```{r}
# Scatter plot of Heartbeat vs Stress Lvevel
plot7 <- ggplot(data, aes(x = Heart.Rate, y = Stress.Level)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "A: Increased stress leads to quicker heartbeats",
       x = "Heartbeat",
       y = "Stress Level")

# Scatter plot of Heartbeat vs Quality of Sleep
plot8 <- ggplot(data, aes(x = Heart.Rate, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "B: Faster pulse, worse quality of sleep",
       x = "Heartbeat",
       y = "Quality of Sleep")

#Show the plots
plot7
plot8
```

```{r}
# Scatter plot of Sleep Quality based on Sleep Quality
plot9 <- ggplot(data, aes(x = Physical.Activity.Level, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "A: Sleep Quality based on Sleep Quality",
       x = "Physical Activity",
       y = "Sleep Quality")

# Scatter plot of Sleep Duration based on Sleep Quality
plot10 <- ggplot(data, aes(x = Sleep.Duration, y = Quality.of.Sleep)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "B: Sleep Duration based on Sleep Quality",
       x = "Heartbeat",
       y = "Quality of Sleep")

#Show the plots
plot9
plot10
```

```{r}
# Scatter plot of Stress Level based on Physical Activity
plot11 <- ggplot(data, aes(x = Stress.Level, y = Physical.Activity.Level)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE) +
  labs(title = "Stress Level based on Physical Activity",
       x = "Physical Activity",
       y = "Sleep Quality")

#Show the plot
plot11
```

\

\










## References

-   https://www.psychiatry.org/patients-families/sleep-disorders/what-are-sleep-disorders#:\~:text=Sleep%20disorders%20(or%20sleep%2Dwake,distress%20and%20impairment%20in%20functioning.

-   https://www.mayoclinic.org/diseases-conditions/insomnia/symptoms-causes/syc-20355167#:\~:text=Insomnia%20is%20a%20common%20sleep,level%20and%20affect%20your%20mood.

-   https://my.clevelandclinic.org/health/diseases/8718-sleep-apnea

-   https://www.kaggle.com/code/alessandrogiacometti/sleep-health-analysis-how-to-sleep-better

-   https://www.salute.gov.it/portale/alleanzaCardioCerebrovascolari/dettaglioSchedeAlleanzaCardioCerebrovascolari.jsp?lingua=italiano&id=18&area=Alleanza%20italiana%20per%20le%20malattie%20cardio-cerebrovascolari&menu=malattie#:\~:text=Quando%20i%20valori%20pressori%20di,si%20parla%20di%20ipertensione%20arteriosa.

## 
